---
title: "191119 - FGF19 project analysis"
output: html_notebook
---

```{r}
sessionInfo()
```

####Install packages:
```{r}
#install.packages("calibrate")
#source("https://bioconductor.org/biocLite.R")
#biocLite("edgeR")
#biocLite("biomaRt")
#biocLite("clusterProfiler")
#biocLite("Rgraphviz")
#biocLite("DOSE")
```

####Load packages:
```{r}
library(edgeR)
library(biomaRt)
library(calibrate)
library(clusterProfiler)
library(org.Hs.eg.db)
library(DOSE)
```

####Load data:
```{r}
counts <- read.delim("~/Desktop/170418_htseq_counts_rvr_Chr1_22", row.names = 1)
```

####Select single sample (SS) data, i.e. all recombinant and all endogenous samples
```{r}
cols_endo=c(1:6)
endo_SS=counts[,cols_endo]

cols_rec=c(7:15)
rec_SS=counts[,cols_rec]

```

####Apply cutoff. Use at least 1 count in 3 samples:
```{r}
rec_counts_wco=rec_SS[(rowSums(rec_SS>0)>=3),]
endo_counts_wco=endo_SS[(rowSums(endo_SS>0)>=3),]
```

#edgeR

####DE analysis with the edgeR package (3 replicates)
```{r}
#First we put the data into a DGEList object:
y_rec=DGEList(rec_counts_wco)
y_endo=DGEList(endo_counts_wco)


#Then we conduct TMM normalization to account for compositional difference between the RNA-seq libraries:
y_rec=calcNormFactors(y_rec)
y_endo=calcNormFactors(y_endo)

#The experimental designs are defined:
patient_rec <- factor(c(16,16,16,224,224,224,226,226,226))
treatment_rec <- factor(c("Ctrl","FGF19","CDCA","Ctrl","FGF19","CDCA","Ctrl","FGF19","CDCA"))
design_rec <- model.matrix(~patient_rec+treatment_rec)
rownames(design_rec) <- colnames(y_rec)
design_rec

patient_endo <- factor(c(16,16,375,375,39,39))
treatment_endo <- factor(c("Ctrl","CDCA","Ctrl","CDCA","Ctrl","CDCA"))
design_endo <- model.matrix(~patient_endo+treatment_endo)
rownames(design_endo) <- colnames(y_endo)
design_endo

#Common dispersions are estimated:
y_rec = estimateGLMCommonDisp(y_rec, design=design_rec, verbose=TRUE)
y_endo = estimateGLMCommonDisp(y_endo, design=design_endo, verbose=TRUE)


#Generalized Linear Models are fitted:
fit_rec = glmFit(y_rec, design_rec)
fit_endo = glmFit(y_endo, design_endo)


#Likelihood ratio tests are conducted:
lrt_rec <- glmLRT(fit_rec)
lrt_endo <- glmLRT(fit_endo)
CDCAvsFGF19 <- glmLRT(fit_rec,coef=5)
CDCAvsCtrl <- glmLRT(fit_rec,coef=4)
CtrlvsFGF19 <- glmLRT(fit_rec,contrast=c(0,0,0,-1,1))

```

####Top DE genes are:
```{r}
topTags(lrt_endo)
print(summary(decideTestsDGE(lrt_endo)))
```
```{r}
topTags(CDCAvsFGF19)
print(summary(decideTestsDGE(CDCAvsFGF19)))
```

```{r}
topTags(CDCAvsCtrl)
print(summary(decideTestsDGE(CDCAvsCtrl)))
```

```{r}
topTags(CtrlvsFGF19)
print(summary(decideTestsDGE(CtrlvsFGF19)))

```

####Add official gene symbols to lists and write the results of the DE analysis to a table:
```{r}
results_endo <- topTags(lrt_endo,n = length(y_endo$counts[,1]))
results_CDCAvsFGF19 <- topTags(CDCAvsFGF19,n = length(y_rec$counts[,1]))
results_CDCAvsCtrl <- topTags(CDCAvsCtrl,n = length(y_rec$counts[,1]))
results_CtrlvsFGF19 <- topTags(CtrlvsFGF19,n = length(y_rec$counts[,1]))

mart <- useMart(biomart="ensembl", dataset = "hsapiens_gene_ensembl",host="www.ensembl.org")
ensembl2name <- getBM(attributes=c("ensembl_gene_id","external_gene_name"),mart=mart)

results_endo <- merge(x=results_endo$table, y=ensembl2name, by.x=0, by.y=1, all.x=TRUE)
results_CDCAvsFGF19 <- merge(x=results_CDCAvsFGF19$table, y=ensembl2name, by.x=0, by.y=1, all.x=TRUE)
results_CDCAvsCtrl <- merge(x=results_CDCAvsCtrl$table, y=ensembl2name, by.x=0, by.y=1, all.x=TRUE)
results_CtrlvsFGF19 <- merge(x=results_CtrlvsFGF19$table, y=ensembl2name, by.x=0, by.y=1, all.x=TRUE)
```
```{r}
write.table(as.matrix(results_endo),file="~/Desktop/191120_endo_DEG",sep="\t")
write.table(as.matrix(results_CDCAvsFGF19),file="~/Desktop/191120_CDCAvsFGF19_DEG",sep="\t")
write.table(as.matrix(results_CDCAvsCtrl),file="~/Desktop/191120_CDCAvsCtrl_DEG",sep="\t")
write.table(as.matrix(results_CtrlvsFGF19),file="~/Desktop/191120_CtrlvsFGF19_DEG",sep="\t")
```

####Write the fitted (TMM) values to a table:
```{r}
fit_endo_name <- merge(x=fit_endo$fitted.values, y=ensembl2name, by.x=0, by.y=1, all.x=TRUE)
write.table(as.matrix(fit_endo_name),file="~/Desktop/191120_endo_TMM",sep="\t")

fit_rec_name <- merge(x=fit_rec$fitted.values, y=ensembl2name, by.x=0, by.y=1, all.x=TRUE)
write.table(as.matrix(fit_rec_name),file="~/Desktop/191120_rec_TMM",sep="\t")
```

```{r}
#Next we library-size normalize the TMM values:
TMM_endo=colSums(fit_endo$fitted.values)
TMM_endo_LibrNorm = sweep(fit_endo$fitted.values, 2, TMM_endo, `/`)*10^6
colSums(TMM_endo_LibrNorm)

TMM_rec=colSums(fit_rec$fitted.values)
TMM_rec_LibrNorm = sweep(fit_rec$fitted.values, 2, TMM_rec, `/`)*10^6
colSums(TMM_rec_LibrNorm)
```

```{r}
##PCA on normalized values for rec FGF19 treatment
par(pty="s") 
PCA_norm_rec <- prcomp(t(log2(TMM_rec_LibrNorm+1)))
Matrix_norm_rec <- summary(PCA_norm_rec)
col.def<-c("black","red","blue", "black","red","blue", "black","red","blue")
colors <- col.def

plot(PCA_norm_rec$x[,1],PCA_norm_rec$x[,2],col=colors,pch=20,xlab=paste0("PC1 (", round(Matrix_norm_rec$importance[2,1]*100), "%)"),ylab=paste0("PC2 (", round(Matrix_norm_rec$importance[2,2]*100),"%)"),cex=3)
text(PCA_norm_rec$x[,1],PCA_norm_rec$x[,2],labels=colnames(TMM_rec_LibrNorm), pos=list(1), cex=1)

plot(PCA_norm_rec$x[,3],PCA_norm_rec$x[,4],col=colors,pch=20,xlab=paste0("PC3 (", round(Matrix_norm_rec$importance[2,3]*100), "%)"),ylab=paste0("PC4 (", round(Matrix_norm_rec$importance[2,4]*100),"%)"),cex=3)
text(PCA_norm_rec$x[,3],PCA_norm_rec$x[,4],labels=colnames(TMM_rec_LibrNorm),pos=list(1), cex=1)

plot(PCA_norm_rec$x[,5],PCA_norm_rec$x[,6],col=colors,pch=20,xlab=paste0("PC5 (", round(Matrix_norm_rec$importance[2,5]*100), "%)"),ylab=paste0("PC6 (", round(Matrix_norm_rec$importance[2,6]*100),"%)"),cex=3)
text(PCA_norm_rec$x[,5],PCA_norm_rec$x[,6],labels=colnames(TMM_rec_LibrNorm),pos=list(1), cex=1)
```

```{r}
##PCA on normalized values for endo FGF19 treatment
par(pty="s") 
PCA_norm_endo <- prcomp(t(log2(TMM_endo_LibrNorm+1)))
Matrix_norm_endo <- summary(PCA_norm_endo)
col.def<-c("darkgreen","gold", "darkgreen","gold", "darkgreen","gold")
colors <- col.def

plot(PCA_norm_endo$x[,1],PCA_norm_endo$x[,2],col=colors,pch=20,xlab=paste0("PC1 (", round(Matrix_norm_endo$importance[2,1]*100), "%)"),ylab=paste0("PC2 (", round(Matrix_norm_endo$importance[2,2]*100),"%)"),cex=3)
text(PCA_norm_endo$x[,1],PCA_norm_endo$x[,2],labels=colnames(TMM_endo_LibrNorm), pos=2, cex=1)

plot(PCA_norm_endo$x[,3],PCA_norm_endo$x[,4],col=colors,pch=20,xlab=paste0("PC3 (", round(Matrix_norm_endo$importance[2,3]*100), "%)"),ylab=paste0("PC4 (", round(Matrix_norm_endo$importance[2,4]*100),"%)"),cex=3)
text(PCA_norm_endo$x[,3],PCA_norm_endo$x[,4],labels=colnames(TMM_endo_LibrNorm),pos=list(1), cex=1)

plot(PCA_norm_endo$x[,5],PCA_norm_endo$x[,6],col=colors,pch=20,xlab=paste0("PC5 (", round(Matrix_norm_endo$importance[2,5]*100), "%)"),ylab=paste0("PC6 (", round(Matrix_norm_endo$importance[2,6]*100),"%)"),cex=3)
text(PCA_norm_endo$x[,5],PCA_norm_endo$x[,6],labels=colnames(TMM_endo_LibrNorm),pos=list(1), cex=1)
```


#Gene Ontology and KEGG Pathway Analysis
####Load the data for GO analysis:
```{r}
#Note: It's better to use ENSEMBL or ENTREZ IDs for the GO analysis. The clusterProfiler package only accepts ENTREZ IDs for KEGG pathway analysis.
CDCA_unique =  read.delim("~/Desktop/191120_CDCA_rec_unique.txt", header=F)
FGF19_unique =  read.delim("~/Desktop/191120_FGF19_rec_unique.txt", header=F)
ENDO = read.delim("~/Desktop/191120_FGF19_endo.txt", header=F)
```

#### Biological Process GO analysis
```{r}
CDCA_BP <- enrichGO(gene = CDCA_unique$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(CDCA_BP))

FGF19_BP <- enrichGO(gene = FGF19_unique$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.10,
                qvalueCutoff  = 0.10,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(FGF19_BP))

ENDO_BP <- enrichGO(gene = ENDO$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.10,
                qvalueCutoff  = 0.10,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(ENDO_BP))
```

```{r}
#Plot top 15 BPs as dotplots
dotplot(CDCA_BP, showCategory=15)
dotplot(FGF19_BP, showCategory=15)
dotplot(ENDO_BP, showCategory=15)
```

```{r}
#Plot top 10 BPs as a cnetplot (GO terms are in brown bubbles, and gene names are in grey bubbles)
cnetplot(CDCA_BP, 10)
#cnetplot(FGF19_BP, 10)
#cnetplot(ENDO_BP, 10)
```

```{r}
#write the analysis results to file:
write.table(as.matrix(summary(CDCA_BP)),file="~/Desktop/191121_Summary_CDCArec_unique_BP.txt",sep="\t")
write.table(as.matrix(summary(FGF19_BP)),file="~/Desktop/191121_Summary_FGF19rec_unique_BP.txt",sep="\t")
write.table(as.matrix(summary(ENDO_BP)),file="~/Desktop/191121_Summary_FGF19endo_BP.txt",sep="\t")
```

#### Molecular Functions GO analysis
```{r}
CDCA_MF <- enrichGO(gene = CDCA_unique$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "MF",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(CDCA_MF))

FGF19_MF <- enrichGO(gene = FGF19_unique$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "MF",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.10,
                qvalueCutoff  = 0.10,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(FGF19_MF))

ENDO_MF <- enrichGO(gene = ENDO$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "MF",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.10,
                qvalueCutoff  = 0.10,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(ENDO_MF))
```

```{r}
#Plot top 15 MFs as dotplots
dotplot(CDCA_MF, showCategory=15)
dotplot(FGF19_MF, showCategory=15)
dotplot(ENDO_MF, showCategory=15)
```

```{r}
#Plot top 10 MFs as a cnetplot (GO terms are in brown bubbles, and gene names are in grey bubbles)
cnetplot(CDCA_MF, 10)
#cnetplot(FGF19_MF, 10)
#cnetplot(ENDO_MF, 10)
```

```{r}
#write the analysis results to file:
write.table(as.matrix(summary(CDCA_MF)),file="~/Desktop/191121_Summary_CDCArec_unique_MF.txt",sep="\t")
write.table(as.matrix(summary(FGF19_MF)),file="~/Desktop/191121_Summary_FGF19rec_unique_MF.txt",sep="\t")
write.table(as.matrix(summary(ENDO_MF)),file="~/Desktop/191121_Summary_FGF19endo_MF.txt",sep="\t")
```

#### Cellular component GO analysis

```{r}
CDCA_CC <- enrichGO(gene = CDCA_unique$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.10,
                qvalueCutoff  = 0.10,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(CDCA_CC))

FGF19_CC <- enrichGO(gene = FGF19_unique$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.10,
                qvalueCutoff  = 0.10,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(FGF19_CC))

ENDO_CC <- enrichGO(gene = ENDO$V1,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.10,
                qvalueCutoff  = 0.10,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(ENDO_CC))
```

```{r}
#Plot top 15 CCs as dotplots
dotplot(CDCA_CC, showCategory=15)
dotplot(FGF19_CC, showCategory=15)
dotplot(ENDO_CC, showCategory=15)
```

```{r}
#Plot top 10 CCs as a cnetplot (GO terms are in brown bubbles, and gene names are in grey bubbles)
#cnetplot(CDCA_CC, 10)
#cnetplot(FGF19_CC, 10)
#cnetplot(ENDO_CC, 10)
```

```{r}
#write the analysis results to file:
write.table(as.matrix(summary(CDCA_CC)),file="~/Desktop/191121_Summary_CDCArec_unique_CC.txt",sep="\t")
write.table(as.matrix(summary(FGF19_CC)),file="~/Desktop/191121_Summary_FGF19rec_unique_CC.txt",sep="\t")
write.table(as.matrix(summary(ENDO_CC)),file="~/Desktop/191121_Summary_FGF19endo_CC.txt",sep="\t")
```

#### KEGG analysis
```{r}
#First change the gene IDs from ENSEMBL to ENTREZ:
CDCA.entrez <- bitr(CDCA_unique$V1, fromType = "ENSEMBL",
        toType = c("ENTREZID"),
        OrgDb = org.Hs.eg.db)

FGF19.entrez <- bitr(FGF19_unique$V1, fromType = "ENSEMBL",
        toType = c("ENTREZID"),
        OrgDb = org.Hs.eg.db)

ENDO.entrez <- bitr(ENDO$V1, fromType = "ENSEMBL",
        toType = c("ENTREZID"),
        OrgDb = org.Hs.eg.db)
```

```{r}
#Then perform the KEGG analysis

#significance cutoff has been loosened due to nothing significant at p/q = 0.05:
CDCA_kegg <- enrichKEGG(gene = CDCA.entrez$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.05)
head(CDCA_kegg)

FGF19_kegg <- enrichKEGG(gene = FGF19.entrez$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.10)
head(FGF19_kegg)

ENDO_kegg <- enrichKEGG(gene = ENDO.entrez$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.10)
head(ENDO_kegg)
```

```{r}
#Plot top 15 CCs as dotplots
dotplot(CDCA_kegg, showCategory=15)
dotplot(FGF19_kegg, showCategory=15)
dotplot(ENDO_kegg, showCategory=15)
```

```{r}
#Plot top 10 KEGGs as a cnetplot (GO terms are in brown bubbles, and gene names are in grey bubbles)
#gene names are ENTREZ IDs here.
cnetplot(CDCA_kegg, 10)
#cnetplot(FGF19_kegg, 10)
#cnetplot(ENDO_kegg, 10)
```

```{r}
#write the analysis results to file:
write.table(as.matrix(summary(CDCA_kegg)),file="~/Desktop/191121_Summary_CDCArec_unique_KEGG.txt",sep="\t")
write.table(as.matrix(summary(FGF19_kegg)),file="~/Desktop/191121_Summary_FGF19rec_unique_KEGG.txt",sep="\t")
write.table(as.matrix(summary(ENDO_kegg)),file="~/Desktop/191121_Summary_FGF19endo_KEGG.txt",sep="\t")
```